---
title: "hw02-CarolineSeip"
author: "Caroline"
date: "September 18, 2018"
output: html_document
---

#Bring regular data in

Start by installing and loading the gapminder dataset and tidyverse. If you already have these installed, just load them using the library function.
```{r}
library(gapminder)
library(tidyverse)
```

#Smell test the data

```{r}
class(gapminder)
```
It is a tibble data frame, a tibble and a data frame

How many columns?
```{r}
ncol(gapminder)
```

How many rows?
```{r}
nrow(gapminder)
```

Or you can use the dim function, which will give you the dimensions of your dataset, telling you the number of rows and columns using only one function
```{r}
dim(gapminder)
```

We can either do it one by one using the class function

```{r}
class(gapminder$country)
```
```{r}
class(gapminder$continent)
```


Or get them all at the same time using the lapply function
```{r}
lapply(gapminder, class)
```

#Explore individual variables

##Quantitative variable: population

To find the range of possible values for a quantitative variable use the range function
```{r}
range(gapminder$pop)
```

##Categoirical variable: continent

To find the possible values for a categorical variable, use the unique function
```{r}
unique(gapminder$continent)
```

To find the typical values (mean, median), spread, distribution use the summary function
```{r}
summary(gapminder)
```

To get these individually you can use the mean function
```{r}
mean(gapminder$lifeExp)
```

To see the spread you can use the range function
```{r}
range(gapminder$lifeExp)
```

Or find the min and max
```{r}
min(gapminder$lifeExp)
max(gapminder$lifeExp)
```

