---
title: "hw02-CarolineSeip"
author: "Caroline"
date: "September 18, 2018"
output: github_document
---

#Bring regular data in

Start by installing and loading the gapminder dataset and tidyverse. If you already have these installed, just load them using the library function.
```{r}
library(gapminder)
library(tidyverse)
```

#Smell test the data

First let's figure out what class the data is, using the 'class' function.
```{r}
class(gapminder)
```
This tells us that the class of the gapminder dataset is a tibble dataframe.

How many columns does the gapminder dataset contain? Use the function 'ncol'
```{r}
ncol(gapminder)
```
The gapminder dataset contains 6 columns.

How many rows does the gapminder dataset contain? Use the function 'nrow'
```{r}
nrow(gapminder)
```
The gapminder dataset contains 1704 rows.

Or you can use the dim function, which will give you the dimensions of your dataset, telling you the number of rows and columns using only one function
```{r}
dim(gapminder)
```
The gapminder dataset contains 1704 rows and 6 columns.

We can find the type of each variable one by one using the 'class' function:
```{r}
class(gapminder$country)
```
```{r}
class(gapminder$gdpPercap)
```

Or we can get them all at the same time using the lapply function:
```{r}
lapply(gapminder, class)
```

#Explore individual variables

##Quantitative variable: population

To find the range of possible values for a quantitative variable, we can use the 'range' function:
```{r}
range(gapminder$pop)
```
This tells us that all values for the quantitative variable 'population' are between 60,011 and 1,318,683,096.

##Categoirical variable: continent

To find the possible values for a categorical variable, we can use the 'unique' function:
```{r}
unique(gapminder$continent)
```
This shows us the 6 different possible values for the categorical variable 'continent'.

To find the typical values (mean, median), spread, and distribution of each variable in a dataset, we can use the 'summary' function:
```{r}
summary(gapminder)
```

To get the typical values for each variable individually we can use the 'mean' function:
```{r}
mean(gapminder$lifeExp)
```
The mean life expectancy is 59.47 years.

To see the spread of each variable individually we can use the 'range' function:
```{r}
range(gapminder$lifeExp)
```
The range of life expectancy is between 23.6 and 82.6 years.

Or we can find the minimum and maximum to show the spread:
```{r}
min(gapminder$lifeExp)
max(gapminder$lifeExp)
```
The minimum life expectancy is 23.6 years and the maximum is 82.6 years.
#Explore various plot types

#Use filter, select and piping

We can also select only certain variables to display or work with, using the 'select' function:
```{r}
select(gapminder, country, year, lifeExp)
```
This selects the variables country, year, and life expectancy.

Filtering
```{r}
filter(gapminder, lifeExp<30)
```

Piping
```{r}
gapminder %>%
  filter(lifeExp<30)
```

